---
title: "Advanced Data Manipulation with dplyr"
---

```{r init, include=F}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE)
.ex <- 1

library(ggplot2)
theme_set(theme_bw() + theme(strip.background = element_blank()))
```

Data analysis involves a large amount of [janitor work](http://www.nytimes.com/2014/08/18/technology/for-big-data-scientists-hurdle-to-insights-is-janitor-work.html) -- munging and cleaning data to facilitate downstream data analysis. This section assumes a [basic familiarity with R](r-basics.html) and [data frames](r-dataframes.html), and demonstrates techniques for advanced data manipulation and analysis with the split-apply-combine strategy. We will use the dplyr package in R to effectively manipulate and conditionally compute summary statistics over subsets of a "big" dataset containing many observations.

## Review

### Our data

We're going to use the yeast gene expression dataset [described on the data frames lesson](r-dataframes.html#our-data). This is a cleaned up version of a gene expression dataset from [Brauer et al. Coordination of Growth Rate, Cell Cycle, Stress Response, and Metabolic Activity in Yeast (2008) _Mol Biol Cell_ 19:352-367](http://www.ncbi.nlm.nih.gov/pubmed/17959824). This data is from a gene expression microarray, and in this paper the authors are examining the relationship between growth rate and gene expression in yeast cultures limited by one of six different nutrients (glucose, leucine, ammonium, sulfate, phosphate, uracil). If you give yeast a rich media loaded with nutrients except restrict the supply of a _single_ nutrient, you can control the growth rate to any rate you choose. By starving yeast of specific nutrients you can find genes that: 

1. **Raise or lower their expression in response to growth rate**. Growth-rate dependent expression patterns can tell us a lot about cell cycle control, and how the cell responds to stress. The authors found that expression of >25% of all yeast genes is linearly correlated with growth rate, independent of the limiting nutrient. They also found that the subset of negatively growth-correlated genes is enriched for peroxisomal functions, and positively correlated genes mainly encode ribosomal functions. 
2. **Respond differently when different nutrients are being limited**. If you see particular genes that respond very differently when a nutrient is sharply restricted, these genes might be involved in the transport or metabolism of that specific nutrient.

You can download the cleaned up version of the data at [bioconnector.org/data](http://bioconnector.org/data). The file is called **brauer2007_tidy.csv**. Later on we'll actually start with the original raw data (minimally processed) and manipulate it so that we can make it more amenable for analysis. 

### Reading in data

We need to load both the dplyr and readr packages for efficiently reading in and displaying this data. We're also going to use many other functions from the dplyr package. Make sure you have these packages installed as described on the [setup page](setup-r.html).

```{r loadpkgs, results='hold'}
# Load packages
library(readr)
library(dplyr)

# Read in data
ydat <- read_csv(file="data/brauer2007_tidy.csv")

# Display the data
ydat

# Optionally, bring up the data in a viewer window
# View(ydat)
```

## The dplyr package

The [dplyr package](https://github.com/hadley/dplyr) is a relatively new R package that makes data manipulation fast and easy. It imports functionality from another package called magrittr that allows you to chain commands together into a pipeline that will completely change the way you write R code such that you're writing code the way you're thinking about the problem.

When you read in data with the readr package (`read_csv()`) and you had the dplyr package loaded already, the data frame takes on this "special" class of data frames called a `tbl`, which you can see with `class(ydat)`. If you have other "regular" data frames in your workspace, the `tbl_df()` function will convert it into the special dplyr `tbl` that displays nicely (e.g.: `iris <- tbl_df(iris)`). You don't have to turn all your data frame objects into tbl_df objects, but it does make working with large datasets a bit easier.

## dplyr verbs

The dplyr package gives you a handful of useful **verbs** for managing data. On their own they don't do anything that base R can't do. Here are some of the _single-table_ verbs we'll be working with in this lesson (single-table meaning that they only work on a single table -- contrast that to _two-table_ verbs used for joining data together, which we'll cover in a later lesson).

1. filter
1. select
1. mutate
1. arrange
1. summarize
1. group_by

They all take a `data.frame` or `tbl_df` as their input for the first argument, and they all return a `data.frame` or `tbl_df` as output.

### filter()

If you want to filter **rows** of the data where some condition is true, use the `filter()` function. 

1. The first argument is the data frame you want to filter, e.g. `filter(mydata, ...`.
2. The second argument is a condition you must satisfy, e.g. `filter(ydat, symbol == "LEU1")`. If you want to satisfy *all* of multiple conditions, you can use the "and" operator, `&`. The "or" operator `|` (the pipe character, usually shift-backslash) will return a subset that meet *any* of the conditions.

- `==`: Equal to
- `!=`: Not equal to
- `>`, `>=`: Greater than, greater than or equal to
- `<`, `<=`: Less than, less than or equal to

Let's try it out. For this to work you have to have already loaded the dplyr package. Let's take a look at [LEU1](http://www.yeastgenome.org/locus/Leu1/overview), a gene involved in leucine synthesis.

```{r filter}
# First, make sure you've loaded the dplyr package
library(dplyr)

# Look at a single gene involved in leucine synthesis pathway
filter(ydat, symbol == "LEU1")

# Optionally, bring that result up in a View window
# View(filter(ydat, symbol == "LEU1"))

# Look at multiple genes
filter(ydat, symbol=="LEU1" | symbol=="ADH2")

# Look at LEU1 expression at a low growth rate due to nutrient depletion
# Notice how LEU1 is highly upregulated when leucine is depleted!
filter(ydat, symbol=="LEU1" & rate==.05)

# But expression goes back down when the growth/nutrient restriction is relaxed
filter(ydat, symbol=="LEU1" & rate==.3)

# Show only stats for LEU1 and Leucine depletion. 
# LEU1 expression starts off high and drops
filter(ydat, symbol=="LEU1" & nutrient=="Leucine")

# What about LEU1 expression with other nutrients being depleted?
filter(ydat, symbol=="LEU1" & nutrient=="Glucose")
```

Let's look at this graphically. Don't worry about what these commands are doing just yet - we'll cover that later on when we talk about ggplot2. Here's I'm taking the filtered dataset containing just expression estimates for LEU1 where I have 36 rows (one for each of 6 nutrients $\times$ 6 growth rates), and I'm _piping_ that dataset to the plotting function, where I'm plotting rate on the x-axis, expression on the y-axis, mapping the value of nutrient to the color, and using a line plot to display the data.

```{r demo1, fig.align='center'}
library(ggplot2)
filter(ydat, symbol=="LEU1") %>% 
  ggplot(aes(rate, expression, colour=nutrient)) + geom_line(lwd=1.5)
```

Look closely at that! LEU1 is _highly expressed_ when starved of leucine because the cell has to synthesize its own! And as the amount of leucine in the environment (the growth _rate_) increases, the cell can worry less about synthesizing leucine, so LEU1 expression goes back down. Consequently the cell can devote more energy into other functions, and we see other genes' expression very slightly raising.

----

**EXERCISE `r .ex``r .ex=.ex+1`**

1. Display the data where the gene ontology biological process (the `bp` variable) is "leucine biosynthesis" (case-sensitive) _and_ the limiting nutrient was Leucine. (Answer should return a 24-by-7 data frame -- 4 genes $\times$ 6 growth rates).
1. Gene/rate combinations had high expression (in the top 1% of expressed genes)? _Hint:_ see `?quantile` and try `quantile(ydat$expression, probs=.99)` to see the expression value which is higher than 99% of all the data, then `filter()` based on that. Try wrapping your answer with a `View()` function so you can see the whole thing. What does it look like those genes are doing? Answer should return a 1971-by-7 data frame.

```{r include=F}
filter(ydat, nutrient=="Leucine" & bp=="leucine biosynthesis")

quantile(ydat$expression, probs=.99)
filter(ydat, expression>2.07)
```

----

### select()

The `filter()` function allows you to return only certain _rows_ matching a condition. The `select()` function returns only certain _columns_. The first argument is the data, and subsequent arguments are the columns you want.

```{r select}
# Select just the symbol and systematic_name
select(ydat, symbol, systematic_name)

# Alternatively, just remove columns. Remove the bp and mf columns.
select(ydat, -bp, -mf)

# Notice that the original data doesn't change!
ydat
```

Notice above how the original data doesn't change. We're selecting out only certain columns of interest and throwing away columns we don't care about. If we wanted to _keep_ this data, we would need to _reassign_ the result of the `select()` operation to a new object. Let's make a new object called `nogo` that does not contain the GO annotations. Notice again how the original data is unchanged.

```{r select2}
# create a new dataset without the go annotations.
nogo <- select(ydat, -bp, -mf)
nogo

# we could filter this new dataset
filter(nogo, symbol=="LEU1" & rate==.05)

# Notice how the original data is unchanged - still have all 7 columns
ydat
```





<!-- ####################################### -->

```{r dplyrlesson, include=F, eval=F}
################################################
library(readr)
library(dplyr)
library(ggplot2)
theme_set(theme_bw())

ydat <- read_csv("data/brauer2007_tidy.csv")
################################################

# select
select(ydat, symbol, systematic_name)
select(ydat, -bp, -mf)
ydat
nogo <- select(ydat, -bp, -mf)
nogo
ydat
filter(nogo, symbol=="LEU1" & rate==.05)

# mutate
# expression is log2 signal/ref (glu-lim at .25h-1)
mutate(nogo, signal=2^expression)
mutate(nogo, signal=2^expression) %>% ggplot(aes(signal, expression)) + geom_point()
# demo (why use log?)
library(tidyr)
mutate(nogo, signal=2^expression)  %>% 
  gather(unit, value, expression, signal) %>% 
  ggplot(aes(value)) + geom_histogram(bins=100) + facet_wrap(~unit, scales="free")
  
# arrange
arrange(ydat, symbol)
arrange(ydat, expression)
arrange(ydat, desc(expression))
filter(ydat, bp=="leucine biosynthesis" & nutrient=="Leucine")
arrange(filter(ydat, bp=="leucine biosynthesis" & nutrient=="Leucine"), symbol)
View(arrange(filter(ydat, bp=="leucine biosynthesis" & nutrient=="Leucine"), symbol))
# demo (leu4 responds to both glucose AND leucine starvartion. why?)
filter(ydat, bp=="leucine biosynthesis") %>% 
  ggplot(aes(rate, expression, col=nutrient)) + geom_line() + facet_wrap(~symbol)

# summarize
summarize(ydat, mean(expression))
summarize(ydat, meanexp=mean(expression))
summarize(ydat, cor(rate, expression))

# group_by
group_by(ydat, nutrient)
group_by(ydat, nutrient, rate)


# combine
summarize(group_by(ydat, nutrient, rate), meanexp=mean(expression))

# demo chain
ydat %>% 
  filter(bp=="leucine biosynthesis") %>%
  group_by(nutrient) %>% 
  summarize(r=cor(rate, expression)) %>% 
  mutate(r=round(r,2)) %>% 
  arrange(r)
```

