---
title: "ggplot2 homework"
output: 
  html_document: 
    keep_md: yes
---
## Key Concepts
> 
- geoms
- aesthetic mappings
- statistical layers
- scales
- ggthemes
- ggsave

```{r init, echo=FALSE, warning = FALSE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=FALSE)
```

## Getting Started

Load the **ggplot2**, **dplyr**, **readr** and **ggthemes** packages, and read the "Brauer" data into R using the `read_csv()` function. 

Make sure you store the results in an object called `ydat`

**n.b. `read_csv()` is *not* the same as `read.csv()`**

```{r loaddata, echo=TRUE}
library(ggplot2)
library(dplyr)
library(readr)
library(ggthemes)

# Preferably: read data from web 
ydat <- read_csv("http://bioconnector.org/data/brauer2007_tidy.csv")

# Alternatively read from file:
# ydat <- read_csv("data/brauer2007_tidy.csv")

# Display the data
ydat
```

## Problem Set

Follow the prompts and use **ggplot2** to reproduce the plots below. 

### Part 1

We can start by taking a look at the distribution of the expression values. 

1) Plot a histogram of the expression variable, and set the bin number equal to 100.

```{r histogram}
ggplot(ydat, aes(x = expression)) +
    geom_histogram(bins = 100) 
```

2) Check the distribution of each nutrient in the data set by adjusting the fill aesthetic. Use the same bin number for this histogram.

```{r histogram_fill}
ggplot(ydat, aes(x = expression, fill = nutrient)) +
    geom_histogram(bins = 100)
```

Wow. That's ugly. Might be a candidate for [accidental aRt](http://accidental-art.tumblr.com/) but not very helpful for our analysis.

3) Now split off the same histogram into a faceted display with 3 columns.

```{r histogram_facet}
ggplot(ydat, aes(x = expression, fill = nutrient)) +
    geom_histogram(bins = 100) +
    facet_wrap(~ nutrient, ncol = 3)
```

The basic exploratory process above confirms that the overall distribution (as well each distribution by nutrient) is normal.

### Part 2

Let's compare the genes with the highest and lowest average expression values. 

We can figure out which these are using some familiar logic:

1. Take the original *ydat* data frame ... 
2. Then *group by* symbol ...
3. Then *summarize* mean of all expression values for that symbol ...
4. Then *arrange* descending by the mean ...
5. Then *filter* for the first or last row.

The code below implements that pipeline in **dplyr** syntax:

```{r, echo=TRUE}
ydat %>%
    group_by(symbol) %>%
    summarise(meanexp = mean(expression)) %>%
    arrange(desc(meanexp)) %>%
    filter(row_number() == 1 | row_number() == n())
```

The output tells us that the gene with the highest mean expression is *HXT3*, while the gene with the lowest mean expression is *HXT6*.

4) Subset the data to only include these genes, and create a stripplot that has expression values as "jittered" points on the y-axis and the gene symbols the x-axis. 

> **HINT** you can add a "jitter" position to `geom_point()` but it's easier to control width of the effect if you use `geom_jitter()`

```{r stripplot}
gene_exp <- 
    ydat %>%
    filter(symbol == "HXT3" | symbol == "HXT6")

ggplot(gene_exp, aes(x = symbol, y = expression)) +
    geom_jitter(width = 0.1)
```

5) Now map each observation to its nutrient by color and adjust the size of the points to be 2.

```{r stripplot_color}

ggplot(gene_exp, aes(x = symbol, y = expression)) +
    geom_jitter(aes(col = nutrient), width = 0.1, size = 2)
```

Although these two genes are on opposite ends of the distribution of average expression values, they both seem to express similar amounts when Glucose is the restricted nutrient. 

### Part 3

Now let's try to make something that has a little bit more of a polished look. 

6) Using **dplyr** logic, create a data frame that has the mean expression values for all combinations of rate and nutrient. Create a plot of this data with rate on the x-axis and mean expression on the y-axis and lines colored by nutrient. 

```{r lineplot}
nutrient_rates <-
    ydat %>%
    group_by(rate, nutrient) %>%
    summarise(meanexp = mean(expression)) 

ggplot(nutrient_rates, aes(x = rate, y =  meanexp)) + geom_line(aes(col=nutrient))

```

7) Add black dotted line (lty=3) that represents the smoothed mean of expression across all combinations of nutrients and rates. 

```{r lineplot_smooth}
ggplot(nutrient_rates, aes(x = rate, y =  meanexp)) + geom_line(aes(col=nutrient), lty=1) + geom_smooth(col = "black", lty=3, se = FALSE) 
```

8) Change the scale to include breaks for *all* of the rates.

> **HINT** The `read_csv()` function read in the rate variable as continuous rather than discrete. There are a few ways to remedy this, but first see if you can set the scale for the x axis variable without changing the dataframe.

```{r lineplot_scale}
ggplot(nutrient_rates, aes(x = rate, y =  meanexp)) + geom_line(aes(col=nutrient), lty=1) + geom_smooth(col = "black", se = FALSE, lty=3) + scale_x_continuous(breaks = nutrient_rates$rate)
```

9) By default `ggplot()` will name the x and y axes with names of their respective variables. You might want to apply more meaningful labels. Change the name of the x-axis to "Rate", the name of the y-axis to "Mean Expression" and the plot title to "Mean Expression By Rate (Brauer)"

> **HINT** `?labs` will pull up the **ggplot2** documentation on axes labels and plot titles.

```{r lineplot_labels}
ggplot(nutrient_rates, aes(x = rate, y =  meanexp)) + geom_line(aes(col=nutrient), lty=1) + geom_smooth(col = "black", se = FALSE, lty=3) + scale_x_continuous(breaks = nutrient_rates$rate) + xlab("Rate") + ylab("Mean Expression") + ggtitle("Mean Expression By Rate (Brauer)")
```

10) Add a theme from the **ggthemes** package. The plot below is based on Edward Tufte's book *The Visual Display of Quantitative Information*. Choose a theme that you like, but choose wisely ... some of these themes will override other adjustments you've made to your plot above, including axis labels.

> **HINT** You can either do this by trial-and-error or check out the package vignette to get an idea of what each theme looks like: <https://github.com/jrnold/ggthemes>

```{r lineplot_final}
ggplot(nutrient_rates, aes(x = rate, y =  meanexp)) + geom_line(aes(col=nutrient), lty=1) + geom_smooth(col = "black", se = FALSE, lty=3) + scale_x_continuous(breaks = nutrient_rates$rate) + xlab("Rate") + ylab("Mean Expression") + ggtitle("Mean Expression By Rate (Brauer)") + theme_tufte()
```

11) The last step is to save the plot you've created. Write your plot to a 10 X 6 PDF using a **ggplot2** function.

```{r, eval = FALSE, echo = FALSE}

p <- ggplot(nutrient_rates, aes(x = rate, y =  meanexp)) + geom_line(aes(col=nutrient), lty=3) + geom_smooth(col = "black", se = FALSE, lty=2) + scale_x_continuous(breaks = nutrient_rates$rate) + ylab("Mean Expression") + ggtitle("Mean Expression By Rate (Brauer)") + theme_tufte()

p %>% ggsave(filename = "plot.pdf", width = 10, height = 6)

```
